Hydrous melting and partitioning in the deep upper mantle: insights from experiments in the system MgO-SiO$_2$-H$_2$O

### 2015-07-07 MSH.tex
Add

\title{Hydrous transition zone melting  
%\title{High-pressure water-rich melting in the system MgO-SiO$_2$-H$_2$O: Insights into hydrous transition zone melting}
%\tnotetext[mytitlenote]{Fully documented templates are available in the elsarticle package on \href{http://www.ctan.org/tex-archive/macros/latex/contrib/elsarticle}{CTAN}.}

@@

Constraining melt compositions remains a major challenge in high-pressure experimental petrology. The primary problem is that hydrous melts are unquenchable, which means that their compositions must be estimated indirectly. Compositions of partial hydrous melts have been estimated using mass balance, diamond traps \citep{MSUP2007}, and the Mg:Si ratio has been estimated by defocusing the beam of an electron probe microanalyser (EPMA) \citep[e.g.][]{YII2004}. Unfortunately, all three of these techniques are accompanied by significant uncertainties, especially in terms of the water content. An alternative technique is to constrain the liquidus temperature at a given composition, thus circumventing the need to estimate water concentrations in the melt after the experiment (when much of the water has been lost). The problem with this technique is that at ambient pressures, hydrous phases are limited in their maximum water content. In the MgO-SiO$_2$ system, the lack of a thermally stable stoichiometric SiO$_2$.H$_2$O compound means that brucite has been used as the water source. This limits water concentrations in model peridotitic systems (Mg$\sim$1.4 Si) to $\sim$18 wt\% H$_2$O. Thus, studies either extrapolate high temperature liquidus curves to high water contents or add water in liquid form. Extrapolation is prone to large errors, and accurately adding liquid water to the small volumes of high pressure capsules is essentially impossible.

@@

Capsules were created from 2 mm-diameter Pt$_{90}$Rh$_{10}$ and Au rods. They were cut by wire saw into 1 mm thick disks, into which were spark-eroded six holes in three rows, 250 microns wide and 700 microns deep. The capsules were cleaned by cycling between an acetone ultrasonic bath (15 minutes) and 1000 $^{\circ}$C furnace (20 minutes) three times. Any remaining contamination was removed with W$_{75}$Re$_{25}$ needle followed by a further trip to the ultrasonic bath. Capsule chambers were filled with powders of different compositions using a W$_{75}$Re$_{25}$ needle. Small pieces of tape were used to cover the other holes to avoid contamination.

@@

\citep{NM2008}
\citep{DM2010}
\citep{ZK1998}

@@

%\citep{SU2001}
%\citep{KOM2005}
%\citep{AA1980}
%\citep{Inoue1994}
%\citep{Wunder1998}
%\citep{AFRH2001}
%\citep{Nagano2002}
%\citep{MSUP2007}
%\citep{MFY2002}
%\citep{Sta2004}
%\citep{MFY2004}
%\citep{SUTG2001}

@@

\cite{DDFK2005}
\citep{LSOK2011}
\citep{OMY2000}

>>>>

\title{Hydrous melting and partitioning in the deep upper mantle: insights from experiments in the system MgO-SiO$_2$-H$_2$O}
%\tnotetext[mytitlenote]{Fully documented templates are available in the elsarticle package on \href{http://www.ctan.org/tex-archive/macros/latex/contrib/elsarticle}{CTAN}.}

@@

Constraining melt compositions remains a major challenge in high-pressure experimental petrology. The primary problem is that hydrous melts are unquenchable, which means that their water concentrations must be estimated indirectly. Compositions of partial hydrous melts have been estimated using a mixture of mass balance and defocused-beam electron probe measurements, which to provide Mg:Si ratios and a rough estimate of water content from deficits in totals. Such measurements are conducted on melt pockets \citep[e.g.][]{OMY2000, DDFK2005, LSOK2011} or melts isolated within diamond traps \citep{MSUP2007}. Unfortunately, these techniques are accompanied by significant uncertainties, especially in terms of water content. An alternative technique is to constrain the liquidus temperature at a given composition, thus circumventing the need to estimate water concentrations in the melt \emph{ex-situ} (when much of the water has been lost). The problem with this technique is that at ambient pressures, hydrous phases are limited in their maximum water content. For example, in the MgO-SiO$_2$ system, maximum water contents in a solid starting material are provided by a mixture of brucite and quartz. Thus, studies either extrapolate high temperature liquidus curves to high water contents or add water in liquid form. Extrapolation is prone to large errors, and accurately adding liquid water to the small volumes of high pressure capsules is essentially impossible.

@@

Capsules were created from 2 mm-diameter Pt$_{90}$Rh$_{10}$ and Au rods, cut by wire saw into 1 mm thick disks. Into each disk were spark-eroded two rows of three holes, each 250 microns in diameter and 700 microns deep. The capsules were cleaned by cycling between an acetone ultrasonic bath (15 minutes) and 1000 $^{\circ}$C furnace (20 minutes) three times. Any remaining contamination was removed with a W$_{75}$Re$_{25}$ needle followed by a further trip to the ultrasonic bath. Capsule chambers were filled with powders of different compositions using a W$_{75}$Re$_{25}$ needle. Small pieces of tape were used to temporarily cover the other holes to avoid contamination.

@@

@@

%\citep{SU2001}
%\citep{KOM2005}
%\citep{AA1980}
%\citep{Wunder1998}
%\citep{AFRH2001}
%\citep{Nagano2002}
%\citep{MSUP2007}
%\citep{MFY2002}
%\citep{Sta2004}
%\citep{MFY2004}
%\citep{SUTG2001}
%\citep{NM2008}
%\citep{DM2010}
%\citep{ZK1998}

@@

Partitioning of hydrogen between mineral and melt phases can be described using a partition coefficient D:
\begin{equation}
  D^{a/b}_{H_2O} = c^a_{H_2O} / c^b_{H_2O}
\end{equation}
where $c^i_{H_2O}$ is the concentration of H$_2$O (in wt\%) in phase $i$ \citep{KB2006}. In quenchable phases, there are a number of ways that concentration can be measured, including FTIR, SIMS and ERDA. In the case of melts, which are unquenchable, $c$ in partitioning studies has most commonly been estimated by mass balance or deficit in microprobe totals. In two studies on Fe-free wadsleyite, \cite{DDFK2005} and  \cite{LSOK2011} present estimates of water content in melts calculated from microprobe deficits. A similar study on ringwoodite was undertaken by \cite{OMY2000}. One striking observation is the extreme differences in water content estimated for melts in equilibrium with wadsleyite at 15 GPa and ringwoodite at 17-20 GPa. For example, melt in equilibrium with wadsleyite at 1400 C was estimated to contain 10.6-13.3 wt wt\% H$_2$O, while melt compositions in equilibrium with ringwoodite suggest water contents of 57-66 wt\% H$_2$O. Modelled Mg$_2$SiO$_4$ melting temperatures increase by only 40 K between 15 and 20 GPa, and entropies of melting at these two pressures are not very different (92.4 vs. 97 J/K/mol), so unless the thermodynamics of mixing in the liquid change markedly between these pressures, estimates of $c^{melt}_{H_2O}$ for the ringwoodite or wadsleyite studies (or both) must be somewhat inaccurate. We note that two of the three values provided by \cite{OMY2000} fall above the temperatures indicating that H$_2$O is the only H-bearing species in the melt \citep{SS1985}, and that the melting temperatures of wadsleyite + 20.4 wt\% H$_2$O (40 mol\% H$_2$O in Figure \ref{fig:partitioning}) in the wadsleyite studies fall below that of the ame composition at 5.5 GPa \citep{Inoue1994}. Both of these interpretations appear unlikely.

Here, we recalculate the wadsleyite/melt and ringwoodite/melt partitioning values with our melt model, assuming that the thermodynamic properties of the melt remain essentially constant between 13 and 20 GPa. Water concentrations in the solid are taken from the original studies. We restrict our analysis to data from 1100--1400 C, where the Mg:Si ratios of the melts are similar to that in forsterite. The average $D^{wad/melt}_{H_2O}$ (0.44) and $D^{rw/melt}_{H_2O}$ values (0.6) produce the lines in Figure \ref{fig:partitioning}. Note that $D^{wad/rw}_{H_2O}$ = 0.72, implying that at equilibrium, the lower transition zone should contain more water than the upper transition zone; the opposite conclusion to that reached by \cite{DDFK2005} using the same data XXXX check this!!. This prediction could explain observations of low velocity layers at ca. 520 km depth \citep{JDH2010}, in an extension of the water filter model of \cite{BK2003}.

An accurate thermodynamic model for hydrous melts can also reveal the mechanism of hydrogen incorporation into nominally anhydrous minerals. This arises from the equilibrium constant for the reaction between oxygen in the silicate, water in the melt, and OH groups in the solid: XXXX correct this!!
\begin{equation}
  K = \frac{a_{OH}^n}{a_{H_2O}^{n/2}a_O^{n/2}}
\end{equation}

Isolated protons (n=0.5), proton pairs (n=1) and proton . Making the approximation that $a_O = 1$, which is probably reasonable at low H concentrations in the solid 
Making the assumption that concentration of H$_2$O in the solid is proportional to the activity of OH in the melt, we therefore have
\begin{equation}
  c \propto a_{H_2O}^n
\end{equation}
where $n$ relates to the mechanisms of hydrogen incorporation into the solid \citep{KB2006}
